---
layout: post
title: "Rips Filtration Tutorial for the R-TDA Package"
author: "Robin Belton and Ben Holmgren"
date: "July 22, 2019"
---

{% include lib/mathjax.html %}

```{r pkgs, echo = FALSE, message = FALSE}
# Initialize Environment
#########################

# Set CRAN location for installing packages
############################################
r <- getOption("repos")
r["CRAN"] <- "http://cran.us.r-project.org"
options(repos = r)
rm(r)

## Helper function to install a CRAN package if not installed and load
######################################################################
use_package <- function(p) {
if (!is.element(p, installed.packages()[,1]))
  install.packages(p, dep = TRUE)
library(p, character.only = TRUE)
}

# Load base R libraries (no need to install)
#############################################
library(grid)

# Check and install CRAN packages
##################################
use_package('png') # For importing pngs
use_package('gridExtra') # For grid arrange
```

Introduction
============
Our ultimate goal is to understand Persistent Homology and be able to compute Persistence
Diagrams using the R-TDA package. In order to get there, we must first understand filtrations.
In this tutorial, we will work with the Vietoris Rips Filtration, or Rips Filtration for short.
Suppose we are given a finite set of points in $$\mathbb{R}^n$$ denoted by $$S$$. The set of points,
$$S$$, is topologically uninteresting. How can we make $$S$$ topologically interesting? Yes, you
guessed it - we can construct a Rips complex from the points in $$S$$!  The Rips Filtration is a
specific nested sequence of Rips Complexes over $$S$$ that can be later used to compute Persistent
Homology.

Objectives
==========
**Make objectives stronger.**

* Gain familiarity on $$n$$-simplices and simplicial complexes.
* Be able to construct a Rips Complex and Rips Filtration from a finite set of points in
$$\mathbb{R}^n$$.
* Learn a few exciting applications of the Rips Filtration.
* Compute a Rips Filtration from a finite set of points in $$\mathbb{R}^n$$ using the R-TDA package.

Theory
======

First we breifly describe the mathematical foundations behind a Rips Filtration including
simplicial complexes, filtrations, and the Rips complex.

### Triangle Appreciation

We construct a Rips Complex from simplices of varying dimensions which are dimensional
generalizations of triangles. More specifically, an *$$n$$-simplex* is the convex hull of $$n+1$$
affinely independent points.

```{r simplices_plot, echo = FALSE, fig.align = 'center'}
fig1 <- rasterGrob(readPNG("images/simplices.png"), interpolate=TRUE)
grid.arrange(fig1, ncol=1)
```

A simplicial complex is a nice way of combining simplices. In particular, a *simplicial complex*
is a finite collection of simplices, $$K$$, such that (1) if $$\sigma \in K$$ and $$\tau\leq \sigma$$,
then $$\tau \in K$$, and (2) if $$\sigma, \sigma'\in K$$, then $$\sigma\cap \sigma'$$ is either empty
or a face of both.

```{r simplicialcomplex, echo = FALSE, fig.align = 'center'}
fig1 <- rasterGrob(readPNG("images/simplicialcomplex.png"), interpolate=TRUE)
grid.arrange(fig1, ncol=1)
```

### Rips Complex

Now we explain how to construct a Rips complex from a finite set of points. Let $$S$$ be finite set
of points in $$\mathbb{R}^n$$. Let $$r\geq 0$$. The Rips complex of $$S$$ and $$r$$ is the abstract
simplicial complex of $$\text{VR}(r)$$, which consists of all subsets of diameter at most $$r$$:

$$ \text{VR}(S, r):=\{\sigma\subset S \mid \text{ diam}(\sigma)\leq r\}. $$

**TODO: Add definitions of ASC and faces.**

Geometrically, this means we consider balls of radius, $$\frac{r}{2}$$, centered at each point in
$$S$$. Whenever $d$ balls have pairwise intersections, we add a $$d-1$$ simplex. For this tutorial,
we will use the standard Euclidan distance to compute a Rips complex. However, one could use any
metric.

**Note:** Many people also define $$\text{VR}(r) :=\{\sigma\subset S \mid \text{diam}(\sigma)\leq 2r\}$$.
However, the algorithms used in the R-TDA package use the first definition.

```{r ripscomplex, echo = FALSE, fig.align = 'center'}
fig1 <- rasterGrob(readPNG("images/ripscomplex.png"), interpolate=TRUE)
grid.arrange(fig1, ncol=1)
```

### Filtrations

A *filtration* of a simplicial complex, $$K$$, is a nested sequence of subcomplexes starting at the
empty set and ending with the full simplicial complex i.e.,

$$\emptyset \subset K_1 \subset K_2 \subset ... \subset K.$$


```{r bobcat, echo = FALSE, fig.align = 'center'}
fig1 <- rasterGrob(readPNG("images/bobcat.png"), interpolate=TRUE)
grid.arrange(fig1, ncol=1)
```

Going back to the Rips complex, we consider $$r$$ to be a free parameter. If we vary $$r$$, we get
different Rips complexes. There is often not a best choice for $$r$$, so why not look at all of
them!? Observe if we increase $$r$$, then we get a family of nested Rips complexes which gives rise
to the *Rips filtration*.

Let's work through an example. Let $$S:=\{(0,0),(1,0),(0,2),(1,2),(3,1)\}\subset \mathbb{R}^2$$. We
want to compute a Rips filtration on $$S$$ for all $$r\geq 0$$. Observe:

* when $$r=1$$, the balls of radius $$\frac{1}{2}$$ centered at $$(0,0)$$ and $$(1,0)$$ intersect which
means we add a 1-simplex between $$(0,0)$$ and $$(1,0)$$. Similarly, we add a 1-simplex between
$$(0,2)$$ and $$(1,2)$$.
* when $$r=2$$, we add a two more 1-simplices between $$(0,0), (0,2)$$, and $$(1,2), (1,0)$$.
* when $$r=\sqrt{5}$$, we add a 3-simplex and a 2-simplex.
* when $$r=\sqrt{10}$$, we add a 4-simplex. **How to draw this? Or should I change the example to four points?**

Here is an illustration of the Rips filtration on $$S$$.

```{r ripsfilt2, echo = FALSE, fig.align = 'center'}
fig1 <- rasterGrob(readPNG("images/ripsfilt2.png"), interpolate=TRUE)
grid.arrange(fig1, ncol=1)
```

Applications of Rips Complexes and Filtrations
==============================================

Now that we have discussed the theory, let's talk about some of the applications! A Rips
filtration is used on point cloud data. A common form of these types of data are location data.
For example:

* Galaxy Data which provides coordinates of stars in different galaxies
* GPS coordinates of Airports on Earth
* GPS coordinates of crimes in the United States

The Rips complex provides information on how close the data points are to each other. The
topological information given by taking a Rips filtration can be studied using Persistent
Homology (see next tutorial) which is often used to analyze these types of data.

Furthermore, Rips complexes are used in shape reconstruction. They are nice to use since these
complexes do not favor a specific type of alignment of the input. **TODO: Add citation to Attali paper**

**Maybe fractal dimension?**

These are just a few of the applications. There are many more!

Computing the Rips Filtration using the R-TDA Package
=====================================================

### Toy Example
First, let's go back to the example where $$S:=\{(0,0),(1,0),(0,2),(1,2),(3,1)\}$$. We will compute the Rips filtration using the R-TDA package.

```{r dataset, echo = TRUE, message = TRUE}
library(TDA) # upload TDA package

S <- cbind(c(0,1,0,1,3),c(0,0,2,2,1)) # write S into R

maxscale <- 5 # limit of the filtration

maxdimension <- 2 # components , loops, and voids

S_RipsFilt <- ripsFiltration(S, maxdimension, maxscale, dist = "euclidean", printProgress = FALSE)
```

**TODO: Add example with a "weird" distance**

### Providence Coffee Shop Location Data
Lastly, we will work through an example with real data!


